Reproducibility of SNV-calling in multiple sequencing runs from single tumors

### 2015-06-04 PAPER/paper.tex
Define SNV on first use.

We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNV-calling. We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates. We further tested whether additional filtering increased or decreased the size of the overlap.  We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966). We further found that using filtering after SNV-calling removed the overlap completely. We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.

@@

However, these and other efforts have produced very few results for patients in terms of improved treatments. It may be that we simply haven't yet analyzed the data in the most informative manner. Another factor may be that the data are noisy. It is well known that sequencing and SNV-calling pipelines are not error-free. For example, different SNV-calling pipelines can return different results on the same data \citep{SNPcall}. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low. Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls. Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue. And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.

>>>>

We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in calling single-nucleotide variants (SNVs). We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates. We further tested whether additional filtering increased or decreased the size of the overlap.  We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966). We further found that using filtering after SNV-calling removed the overlap completely. We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.

@@

However, these and other efforts have produced very few results for patients in terms of improved treatments. It may be that we simply haven't yet analyzed the data in the most informative manner. Another factor may be that the data are noisy. It is well known that sequencing and variant-calling pipelines are not error-free. For example, different pipelines for calling single-nucleotide variants (SNVs) can return different results on the same data \citep{SNPcall}. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low. Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls. Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue. And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.
